Eviction and Crime: Evidence from Boston

### 2023-04-12 writing/paper/main.tex
Updates from Overleaf

\begin{titlepage}
\title{Eviction and Crime: Evidence from Boston}
\author{Arjun Shanmugam}
\date{\today}
\maketitle
\begin{abstract}
\noindent This paper investigates the effects of eviction on crime incident frequency within 500 meters of the property where the eviction occurred. I use a difference-in-differences research design based on the staggered conclusion of eviction cases in Boston, Massachusetts between June 2019 and November 2022. Using properties where eviction was unsuccessful as the counterfactual for properties where eviction was successful, I find negative effects of eviction on crime which persist for about eleven months after case resolution. The magnitudes of these effects are approximately 0.5 percent of the mean number of crimes per month within 500 meters of a property in 2017. Placebo tests using larceny and burglary incident responses as an outcome variable suggest that these effects are driven by the movement of individuals involved in crime incidents outside the radii of properties where evictions are successful.

@@

It is, however, possible that the assumptions underlying the difference-in-differences framework hold only conditional on certain pre-treatment characteristics. I algorithmically select pre-treatment characteristics which are significant predictors of trends in the outcome variable over the treatment period. I estimate a propensity score model which models the probability of a successful eviction based on the selected pre-treatment characteristics. I show that reweighting counterfactual properties using inverse propensity scores eliminates significant differences in observable characteristics between properties where evictions were successful and properties where evictions were unsuccessful. Then, I produce new difference-in-differences estimates using \cite{santanna_doubly_2020}'s doubly robust estimator, controlling for the selected pre-treatmennt characteristics. It models outcome differences and rewights counterfactual properties by their inverse propensity for successful eviction \citep{santanna_doubly_2020}.

Doubly robust estimates suggest that eviction has small negative effects on the prevalence of crime incidents in the 500 meters surrounding a property. These estimates are significantly different from zero for most of the first 10 months after case resolution. Their magnitudes are approximately 0.5 percent of the number of crime incident responses within 500 meters of the average property in my sample during 2017. I do not find evidence of pre-treatment differences in crime incident response trends between properties where evictions were successful and properties where evictions were unsuccessful. I hypothesize that these results are driven by the movement of individuals involved in crime incidents outside the immediate radii of properties where evictions are successful. To test this hypothesis, I produce ``placebo'' estimates of the effect of eviction on larceny and burglary crime incident responses. Individuals who commit larceny and burglary generally do not do so within 500 meters of their own homes \textcolor{red}{CITE THIS}. Thus, if my estimates of the effect of the eviction on crime really are driven by the forced removal of individuals involved in crime from a property's immediate radius, then we would not expect to measure an effect of eviction on larceny and burglary crime incident responses. Indeed, these estimates are indistinguishable from 0, supporting my theory about the mechanism via which eviction at a property reduces crime incident responses in its immediate vicinity.

@@

Section 2 discusses the institutional context of the study. Section 3 discusses in greater depth the data I obtain and the dataset I assemble for my analysis. Section 4 outlinnes my empirical strategy. Section 5 provides and discusses results, and Section 6 concludes.

\section{Institutional Context}
    \subsection{Eviction in Massachusetts}
        \subsubsection{The Massachusetts Housing Court}
        \subsubsection{The Eviction Process}
    \subsection{Crime Incident Responses}

>>>>

\begin{titlepage}
\title{Eviction and Crime: Evidence from Boston}
\author{Arjun Shanmugam}
\date{\today}
\maketitle
\begin{abstract}
\noindent This paper investigates the effects of eviction on the frequency of crime incident responses in the immediate vicinity of a property. I use a difference-in-differences research design based on the staggered conclusion of eviction cases in Boston, Massachusetts between June 2019 and March 2020. Using properties where eviction was unsuccessful as the counterfactual for properties where eviction was successful, I find negative effects of eviction on crime incident responses which persist for about eleven months after case resolution. The magnitudes of these effects are approximately 0.5 percent of the mean number of crimes per month in the immediate vicinity of a property in 2017. Placebo tests using larceny and burglary incident responses as an outcome variable suggest that these effects are driven by the movement of individuals involved in crime incidents outside the radii of properties where evictions are successful. Analysis of treatment effects using only cases resolved before the beginning of the pandemic provides evidence against the hypothesis that my results are driven by a pandemic-related shift in characteristics of evicted tenants.

@@

It is, however, possible that the assumptions underlying the difference-in-differences framework hold only conditional on certain pre-treatment characteristics. I algorithmically select pre-treatment characteristics which are significant predictors of trends in the outcome variable over the treatment period. I estimate a propensity score model which models the probability of a successful eviction based on the selected pre-treatment characteristics. I show that reweighting counterfactual properties using inverse propensity scores eliminates significant differences in observable characteristics between properties where evictions were successful and properties where evictions were unsuccessful. Then, I produce new difference-in-differences estimates using \cite{santanna_doubly_2020}'s doubly robust estimator, controlling for the selected pre-treatmennt characteristics. It models outcome differences and rewights counterfactual properties by their inverse propensity for being the site of a successful eviction \citep{santanna_doubly_2020}.

Doubly robust estimates suggest that eviction has small negative effects on the prevalence of crime incidents in the 500 meters surrounding a property. These estimates are significantly different from zero for most of the first 10 months after case resolution. Their magnitudes are approximately 0.5 percent of the number of crime incident responses within 500 meters of the average property in my sample during 2017. I do not find evidence of pre-treatment differences in crime incident response trends between properties where evictions were successful and properties where evictions were unsuccessful \textcolor{red}{make sure this is true}.

I hypothesize that these results are driven by the movement of individuals involved in crime incidents outside the immediate radii of properties where evictions are successful. To test this hypothesis, I produce ``placebo'' estimates of the effect of eviction on larceny and burglary crime incident responses. Individuals who commit larceny and burglary generally do not do so within 500 meters of their own homes \textcolor{red}{CITE THIS}. Thus, if my estimates of the effect of the eviction on crime really are driven by the forced removal of individuals involved in crime from a property's immediate radius, then we would not expect to measure an effect of eviction on larceny and burglary crime incident responses. Indeed, these estimates are indistinguishable from 0, supporting my theory about the mechanism via which eviction at a property reduces crime incident responses in its immediate vicinity.

@@

Section 2 discusses the institutional context of the study. Section 3 discusses in greater depth the data I obtain and the dataset I assemble for my analysis. Section 4 outlines my empirical strategy. Section 5 provides and discusses results, and Section 6 concludes.

ADD DISCUSISON NOF PANDEMIC: ONLY THE MOST SERIOUS EVICTIONS PROCEEDED, BIASING EVICTEES TO BE MORE SERIOUS OFFENDERS

\section{Institutional Context}
    \subsection{Eviction in Massachusetts}
        \subsubsection{Legal Landscape of Eviction}
        Eviction cases—known formally as summary process cases—fall under the purview of the Massachusetts Trial Court. In particular, three sub-departments of the Trial Court have jurisdiction over summary process cases: District Court, Boston Municipal Court, and Housing Court \textcolor{red}{CITE}. Evictions may be filed in any of these three courts. The District Court maintains a large number of locations throughout the state of Massachusetts and Boston Municipal Court operates only within the city limits of Boston.

        The vast majority of summary process cases are adjudicated in the Housing Court, established in 1971 with the specific purpose of dealing with housing-related matters. The court has been expanded several times since then. Since the passage of the most recent Housing Court expansion law in 2017, every Massachusetts resident has access to a Housing Court \citep{gee_housing_2017}. Across the state, 15 judges preside over cases across six divisions: Central, Eastern, Metro South, Northeast, Southeast, and Western.

        \begin{figure}
            \centering
            \includegraphics{research/institutional_context/sources/map_of_where_court_sits_-_jan_2019.pdf}
            \caption{Divisions of the Massachusetts Housing Court}
        \end{figure}

        Four features distinguish the Housing Court as a legal institution from the District Court and Boston Municipal Court \citep{noauthor_housing_nodate}. First, it is led by justices with significant knowledge and experience when it comes to housing-related legal matters, such as summary process cases. Second, it is staffed by housing specialists, employees of the Court with detailed knowledge of Massachusetts housing law who provide information and referrals to resources for landlords and tenants. Third, it offers a service known as mediation, in which cases may be resolved prior to arguments in front of a judge. Mediation is facilitated by a housing specialist, who helps the defendant and plaintiff come to an agreement and records promises made by both sides. Rather than risk defeat at the hands of a judge, many tenants and landlords prefer to reach mutually agreeable terms of resolution during mediation. If either party violates the terms of the specified mediation agreement, the other may return to the judge in a more favorable legal position. Fourth, either party in a summary process case filed in District Court or Boston Municipal Court has the right to transfer the case to the Housing Court at any time prior to trial . For these reasons, tenant advocacy groups and landlord associations alike in Massachusetts generally recommend filing summary process cases in the Housing Court \citep{noauthor_massachusetts_nodate, noauthor_housing_nodate}.

        \subsubsection{The Eviction Process}
        Tenants in Massachusetts can rent from landlords under the terms of a lease or without one. If a tenant is renting from a landlord without a lease, their tenancy is \textit{at will}. The nature of the agreement between the tenant and the landlord—the \textit{landlord-tenant agreement}—determines the steps that the landlord must take to evict the tenant.

        A landlord in Massachusetts begins the eviction process by serving her tenant with a \textit{notice to quit}, which serves as written notice of her desire to terminate tenancy. While a landlord may deliver the notice to quit herself, many landlords choose to hire a local law enforcement officer to do so for around 40 dollars. This guarantees that an independent third party can vouch that the notice was properly served. A notice to quit is served for \textit{nonpayment of rent}, \textit{cause}, or \textit{no fault} and gives an amount of time after which the landlord-tenant agreement will be terminated if no action is taken by the tenant.

        If the notice to quit is served for nonpayment of rent, then regardless of the nature of the landlord-tenant agreement, this time period given by the notice to quit is 14 days. The tenant may \textit{cure} the nonpayment of rent, nullifying the notice to quit, by paying the landlord all owed rent with interest and costs within 14 days \footnote{Tenants at-will do not have the option to cure nonpayment of rent if they have received a separate notice to quit for nonpayment of rent during the last 12 months.}. If the notice to quit is served for cause and a lease defines the landlord-tenant agreement, then the amount of time that given by the notice to quit must be specified in the lease. If the landlord-tenant agreement is at will, then the time period given by the notice to quit is 30 days. No fault notices to quit waiting period of 30 days or one full rental period—whichever is longer—before tenancy can be terminated.

        After the length of time specified by the notice to quit has passed \footnote{If the notice to quit was for nonpayment of rent, the specified length of time must pass without curing of the nonpayment by the tenant.}, the tenant's rental agreement has ended. To continue with the eviction, the landlord is required to hire a local law enforcement officer to serve the tenant with a \textit{summary process summons and complaint}. Massachusetts law says that the summary process summons and complaint must then be filed with the court on a Monday between seven and 30 days after it was served. The date on which the summons and complaint is filed with the court is known as the \textit{entry date}. If a landlord fails to follow any of the above protocol—say, by serving a notice to quit that specifies too short of a time period or by failing to deliver a summary process summons and complaint to the tenant—her case may be dismissed, automatically awarding victory to the tenant.

    \subsection{Crime Incident Responses}
